Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-03-29 paper/AlignedIsotropicGW.tex

As discussed above, BBHs can form as the end point of isolated binary evolution or dynamically in dense stellar environments. The spin misalignments of BBHs is a signature of their formation mechanism; binaries formed through isolated binary evolution are expected to have their spins aligned with the orbital angular momentum, whilst those formed dynamically are expected to have an isotropic distribution.

Rate uncertainties for both of these formation scenarios are large enough that we do not know a priori which will dominate the LIGO detection rate. In the previous section we have shown that a model with isotropic spin directions and 'decreasing' spin magnitudes is favoured by the data.

In this section we assume that the true distribution of BBH spin-orbit misalignments observed by LIGO is a mixture of binaries formed through isolated binary evolution (with aligned spins) and dynamical formation (with isotropic spins).

%
\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{../plots/posterior_on_isotropic_fraction.png}
\caption{\textbf{Fraction of the BBH population coming from an isotropic distribution} The dotted (blue) line shows the flat prior on the fraction of BBHs coming from an isotropic distribution $f_i$. The solid (red) line shows the posterior after O1, assuming that all BHs have their spin magnitude drawn from a uniform distribution. The dashed line assumes a `increasing' distribution $p(a) = 2a$ for BH spin magnitudes, whilst the dot-dash line assumes a `decreasing' distribution $p(a) = 2(1-a)$. We see that regardless of our assumption regarding BH spin magnitudes, there is a preference for a large fraction coming from an isotropic distribution.}
\label{fig:mixture_fraction_posterior}
\end{figure}
%

@@

We fit a mixture model (labelled model 'M') where a fraction $f_i$ ($f_i=\lambda$ in the derivations above) of BBHs have spins drawn from an isotropic distribution, whilst a fraction $1 - f_i$ have their spins aligned with the orbital angular momentum. We assume a flat prior on the fraction $f_i$. To test the robustness of our result, we vary the distribution we assume for BH spin magnitude distributions as with the aligned and isotropic models. We use the `uniform', `increasing' and `decreasing' distributions introduced in Section X, assuming all BHs have their spin magnitude drawn from the same distribution. We calculate the posterior given by Equation \ref{eq:magnitude-dists} plot the posterior distribution/cumulative posterior in Figure~\ref{fig:mixture_fraction_posterior}. We find the mean fraction of BBHs coming from an isotropic distribution is 0.63, 0.71 and 0.78 assuming the `decreasing', `uniform' and `increasing' distributions for spin magnitudes respectively, compared to the prior mean of 0.5. The lower 90\% limits are 0.26, 0.39 and 0.52 respectively, compared to the prior of 0.1. Thus, irrespective of our ignorance of the BH spin magnitude distribution, we find that the current O1 LIGO observations constrain the majority of BBHs to have their spins drawn from an isotropic distribution.

If we assume that the subpopulation of BBHs with isotropically distributed spin orientations corresponds to a subpopulation of dynamically formed BBHs, our result suggests that the majority of merging BBHs observed by LIGO are formed dynamically, rather than through isolated binary evolution.

The evidence ratios of these models to the isotropic distribution with uniform spin magnitudes are 0.85, 0.39 and 0.19. Thus we cannot rule out a mixture with the current data.

@@

The components of this formula are
\begin{itemize}
\item The GW (marginal) likelihood, $p\left(d \mid \chieff\right)$.
  Here we use ``marginal'' because we are (implicitly) integrating
  over all parameters of the signal but $\chieff$.  Note that it is
  the likelihood rather than the posterior that matters for the
  hierarchical analysis; if we are given posterior distributions or
  posterior samples, we need to re-weight to "remove" the prior and
  obtain the likelihood.
\item The population distribution for $\chieff$,
  $p\left( \chieff \mid \lambda \right)$.  This function can be
  parameterised by population-level parameters, $\lambda$.  (In the
  cases discussed above, there are no parameters for the population.)
\item The prior on the population-level parameters, $p(\lambda)$.
\end{itemize}
If we do not care about the individual event $\chieff$ parameters, we
can integrate them out, obtaining
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \int \dd \chieff^i \, p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}
If we are given posterior samples of $\chieff^{ij}$ ($i$ labels the
event, $j$ labels the particular posterior sample) drawn from an
analysis using a prior $p\left( \chieff \right)$, then we can
approximate the integral by a re-weighted average of the population
distribution over the samples (here $p\left( \chieff^{ij} \right)$ is
the prior used to produce the posterior samples):
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \frac{1}{N_i} \sum_{j=1}^{N_i} \frac{p\left( \chieff^{ij} \mid \lambda \right)}{p\left( \chieff^{ij} \right)} \right] p\left(\lambda\right).
\end{equation}

>>>>

%As discussed above, BBHs can form as the end point of isolated binary evolution or dynamically in dense stellar environments. The spin misalignments of BBHs is a signature of their formation mechanism; binaries formed through isolated binary evolution are expected to have their spins aligned with the orbital angular momentum, whilst those formed dynamically are expected to have an isotropic distribution.
%
%Rate uncertainties for both of these formation scenarios are large enough that we do not know a priori which will dominate the LIGO detection rate. In the previous section we have shown that a model with isotropic spin directions and 'decreasing' spin magnitudes is favoured by the data.
%
%In this section we assume that the true distribution of BBH spin-orbit misalignments observed by LIGO is a mixture of binaries formed through isolated binary evolution (with aligned spins) and dynamical formation (with isotropic spins).

In this section we assume that the true distribution of
\acp{BBH} observed by LIGO is a mixture of binaries formed
through isolated binary evolution (with aligned spins) and
dynamical formation (with isotropic spins). Rate uncertainties
for both of these formation scenarios are large enough that we
do not know a priori which will dominate the LIGO detection
rate.

%
\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{../plots/posterior_on_isotropic_fraction.png}
\caption{\textbf{Fraction of the BBH population coming from an
isotropic distribution} The dotted line shows the flat
prior on the fraction of BBHs coming from an isotropic
distribution $f_i$ under the mixture model. The 3 red lines show
the posterior on $f_i$ after O1 under with our various
assumptions regarding BH spin magnitudes.
The solid line shows the posterior assuming that all BHs have
their spin magnitude drawn
from the ``flat'' distribution. The dashed line assumes the
``high'' BH spin magnitude distribution $p(a) = 2a$. The
dot-dash line assumes the ``low'' distribution $p(a) = 2(1-a)$.
We see that for a wide range of assumptions regarding BH spin
magnitudes, the fraction coming from an isotropic distribution
$f_i$ peaks at 1.}
\label{fig:mixture_fraction_posterior}
\end{figure}
%

@@

Following \citet{Stevenson:2017spin},  we fit a mixture model (labelled model 'M' in Figure~\ref{fig:O1-odds}) where a fraction $f_i$ of BBHs have spins drawn from an isotropic distribution, whilst a fraction $1 - f_i$ have their spins aligned with the orbital angular momentum. We assume a flat prior on the fraction $f_i$. To test the robustness of our result, we vary the distribution we assume for BH spin magnitude distributions as with the aligned and isotropic models. We use the ``flat'', ``high'' and ``low'' distributions (Equation~\ref{eq:magnitude-dists}), assuming all BHs have their spin magnitude drawn from the same distribution for both the aligned and isotropic populations. We calculate and plot the posterior on $f_i$ given by Equation~\ref{eq:hierarchical-posterior} ($f_i=\lambda$ in the derivation) in Figure~\ref{fig:mixture_fraction_posterior}.

We find the mean fraction of BBHs coming from an isotropic distribution is 0.63, 0.71 and 0.78 assuming the ``low'', ``flat'' and ``high'' distributions for spin magnitudes respectively, compared to the prior mean of 0.5. The lower 90\% limits are 0.26, 0.39 and 0.52 respectively, compared to the prior of 0.1. In all cases, the posterior peaks at $f_i = 1$. Thus, for a range of assumptions regarding the the BH spin magnitude distribution, we find that the current O1 LIGO observations constrain the majority of BBHs to have their spins drawn from an isotropic distribution.

The evidence ratios of these models to the isotropic distribution with ``flat'' spin magnitudes are 0.85, 0.39 and 0.19 for the ``low'', ``flat'' and ``high'' spin magnitude models. Thus we cannot rule out a mixture with the current data.

@@

The components of this formula are
\begin{itemize}
\item The GW (marginal) likelihood, $p\left(d \mid \chieff\right)$.
  Here we use ``marginal'' because we are (implicitly) integrating
  over all parameters of the signal but $\chieff$.  Note that it is
  the likelihood rather than the posterior that matters for the
  hierarchical analysis; if we are given posterior distributions or
  posterior samples, we need to re-weight to "remove" the prior and
  obtain the likelihood.
\item The population distribution for $\chieff$,
  $p\left( \chieff \mid \lambda \right)$.  This function can be
  parameterised by population-level parameters, $\lambda$.  (In the
  cases discussed above, there are no parameters for the population.)
\item The prior on the population-level parameters, $p(\lambda)$.
\end{itemize}
If we do not care about the individual event $\chieff$ parameters, we
can integrate them out, obtaining
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \int \dd \chieff^i \, p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}
If we are given posterior samples of $\chieff^{ij}$ ($i$ labels the
event, $j$ labels the particular posterior sample) drawn from an
analysis using a prior $p\left( \chieff \right)$, then we can
approximate the integral by a re-weighted average of the population
distribution over the samples (here $p\left( \chieff^{ij} \right)$ is
the prior used to produce the posterior samples):
\begin{equation}
  p\left( \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} \frac{1}{N_i} \sum_{j=1}^{N_i} \frac{p\left( \chieff^{ij} \mid \lambda \right)}{p\left( \chieff^{ij} \right)} \right] p\left(\lambda\right).
  \label{eq:hierarchical-posterior}
\end{equation}
